### 2020-03-27 CombaWeiskopf2020/paper.tex
add sources, reformat for IEEEtran
files provided by Joao and Daniel in a zip, with manuscript source prepared using old template; changed to use IEEEtran (some errors could remain)

\documentclass[10pt,journal,compsoc]{IEEEtran}

\begin{document}

\title{The past and future of visualization for computing in science and engineering}

\author{%
	Joao Comba, Daniel Weiskopf
	\IEEEcompsocitemizethanks{
	\IEEEcompsocthanksitem Joao Comba is at Universidade Federal do Rio Grande do Sul, Brazil\protect\\ 
	\IEEEcompsocthanksitem Daniel Weiskopf is at University of Stuttgart, Germany.
}

\IEEEtitleabstractindextext{%
\begin{abstract}
We provide a retrospective of the Visualization Corner and how its articles covered visualization topics since 2011. This leads us to identify current and future developments in visualization research and their impact on computing in science and engineering.
\end{abstract}
}

% make the title area
\maketitle

The Visualization Corner (VizCorner) is a long-standing department in IEEE Computing in Science \& Engineering (CiSE). It reflects the long-term and fruitful overlap between computer-based visualization and computational science and engineering. As early as 1987, the Workshop on Visualization in Scientific Computing identified the need for visualization for the growing area of simulation-oriented science~\cite{NSFreport1987}. In general, visualization has been playing a great role in visual data analysis and communication in science, engineering, and beyond.

Let us first focus on the scope of the VizCorner by looking into what we have covered there in the last decade. Then we will relate this to current trends in visualization research, leading to possible future developments for visualization in science and engineering.

\section{Retrospective of the VizCorner}

The VizCorner has three column articles per year on average. Our goal is to cover the breadth of visualization research and practice relevant to the readers of this magazine. Therefore, the VizCorner wants to provide a glimpse of basic visualization research questions but also illustrate and highlight the practical impact of visualization. We have been soliciting papers mostly from the visualization community in order to bring in their perspective to the various areas of computational science and engineering.

Let us organize the papers from 2011 until now into themes, indicating the focal areas of the VizCorner. 
We will provide examples of previous papers that illustrate the breadth of visualization possibilities.

\subsubsection{Multidimensional and Multivariate Data Visualization}
% 4 papers:
% Paulovich, F., Silva, C., & Nonato, L. (2012). User-centered multidimensional projection techniques. Computing in Science and Engineering, 14(4), 74-81.
% Da Silva, R., Rauber, P., & Telea, A. (1. 9 2016). Beyond the third dimension: Visualizing high-dimensional data with projections. Computing in Science and Engineering, 18(5), 98-107.
% Doleisch, H., & Hauser, H. (3 2012). Interactive visual exploration and analysis of multivariate simulation data. Computing in Science and Engineering, 14(2), 70-76.
% Heinrich, J., & Weiskopf, D. (2015). Parallel coordinates for multidimensional data visualization: Basic concepts. Computing in Science and Engineering, 17(3), 70-76.

Multidimensional and multivariate data is almost omnipresent in computational science and engineering. Accordingly, we had several papers addressing this topic.
The paper by Helmut Doleisch and Helwig Hauser~\cite{Doleisch2012} nicely shows how the combination of 2D and 3D visualization, flexible degree-of-interest definition, and interactive exploration leads to an effective visual analysis of complex simulation data. They discuss, for example, how multiple simulation attributes from computational fluid dynamics can be visually analyzed. Figure~\ref{fig:simvis} shows a screenshot from their software system, illustrating how different visualization views are integrated in order to support a user-oriented exploration and analysis process.

\begin{figure}[H]
    \begin{center}    
        \includegraphics[width=0.9\textwidth]{simvis.png}
        \caption{Example of multivariate data visualization with the SimVis system~\cite{Doleisch2012}. This screenshot shows the windflow around a building, where turbulent parts are emphasized using a vortex detection method. Image: \copyright{} 2012 IEEE, reprinted from Doleisch and Hauser~\cite{Doleisch2012}.
  \label{fig:simvis}}        
    \end{center}
\end{figure}

Other papers for this theme covered multidimensional projection techniques that reduce multidimensional data to 2D data for visualization purposes. One paper specifically included user-centered control into the projection techniques. Another paper followed a different approach, employing parallel coordinates to show multiple data dimensions simultaneously. Overall, the papers in this category come with quite generic visualizations that can be useful in many areas of science and engineering.

\subsubsection{Flow Visualization}
% 3+1 papers:
%Netzel, R., & Weiskopf, D. (2013). Texture-based flow visualization. Computing in Science and Engineering, 15(6), 96-102.
%Karch, G., Sadlo, F., Weiskopf, D., Hansen, C., Li, G., & Ertl, T. (2012). Dye-based flow visualization. Computing in Science and Engineering, 14(6), 80-86.
%Weinkauf, T., & Theisel, H. (2012). Flow visualization and analysis using streak and time lines. Computing in Science and Engineering, 14(5), 78-84.
% also in HCI: Englund, R., Palmerius, K., Hotz, I., & Ynnerman, A. (1. 5 2018). Touching Data: Enhancing Visual Exploration of Flow Data with Haptics. Computing in Science and Engineering, 20(3), 89-100.

This topic area is a classic one in scientific visualization, considering input from flow simulation and experiments alike. We pick one of the VizCorner papers as an example: Grzegorz Karch and colleagues~\cite{Karch2012} described methods for high-quality computational flow visualization utilizing the metaphor of dye advection. Experimental flow visualization has a long tradition of dye-based observations. With its computational counterpart, we can use the same idea now in a computer-based visualization, see Figure~\ref{fig:karman} as a typical example.

\begin{figure}[th]
	\begin{center}	
		\includegraphics[width=0.9\textwidth]{karman_weno3.png}
		\caption{Computational dye advection~\cite{Karch2012}, applied to a 2D simulation of a von K{\'a}rm{\'a}n vortex street. Image: \copyright{} 2012 IEEE, reprinted from Karch et al.~\cite{Karch2012}.
  \label{fig:karman}}		
	\end{center}
\end{figure}

Other papers described alternative visual representations of flow, for example, by using textures or streak and timelines. Another paper highlighted interaction techniques with haptics that let users virtually touch the data.

\subsubsection{Human-Computer Interaction and Storytelling}

%User perspective, user studies, HCI, perception (4)
%Kurzhals, K., Burch, M., Pfeiffer, T., & Weiskopf, D. (1. 9 2015). Eye Tracking in Computer-Based Visualization. Computing in Science and Engineering, 17(5), 64-71.
%Oliveira, M. (9 2013). Towards more accessible visualizations for color-vision-deficient individuals. Computing in Science and Engineering, 15(5), 80-87.
%Englund, R., Palmerius, K., Hotz, I., & Ynnerman, A. (1. 5 2018). Touching Data: Enhancing Visual Exploration of Flow Data with Haptics. Computing in Science and Engineering, 20(3), 89-100.
%Sauer, F., Neuroth, T., Chu, J., & Ma, K. (1. 11 2016). Audience-Targeted Design Considerations for Effective Scientific Storytelling. Computing in Science and Engineering, 18(6), 68-76.

The latter paper leads us to the next theme that focuses on the human-centered aspects of visualization. In general, interaction plays an essential role in visualization; for example, in order to support scientists in exploring their simulation data. Other VizCorner articles highlighted new developments in eye tracking for interactive visualization or the role of human visual perception. Finally, Franz Sauer and colleagues~\cite{Sauer2016} went all the way to discussing visualization as a means of storytelling that can address a wide range of audiences, from domain scientists to general scientists to the public. For the latter one, they showcased an example of an interactive museum exhibit that visualized plankton populations to museum visitors, see Figure~\ref{fig:museum}.

\begin{figure}[th]
    \begin{center}    
        \includegraphics[width=0.55\textwidth]{museum1.png} \hspace{-3em} \raisebox{12ex}{\includegraphics[width=0.5\textwidth]{museum2.png}}
        \caption{Interactive table with a plankton visualization. Image: \copyright{} 2016 IEEE, reprinted from Sauer et al.~\cite{Sauer2016}. TODO for layout: could you place the right image slightly overlapping on the left on? As to imply something similar to an inset?
  \label{fig:museum}}        
    \end{center}
\end{figure}

\subsubsection{Astronomy and Astrophysics}

%Wenger, S., Ament, M., Steffen, W., Koning, N., Weiskopf, D., & Magnor, M. (5 2012). Interactive visualization and simulation of astronomical nebulae. Computing in Science and Engineering, 14(3), 78-87.
%Aragon-Calvo, M., & Subbarao, M. (1. 11 2015). A Flight through the Universe. Computing in Science and Engineering, 17(6), 96-102.
%Mller, T., & Weiskopf, D. (7 2011). Special-relativistic visualization. Computing in Science and Engineering, 13(4), 85-93.
%Muller, T., & Weiskopf, D. (11 2011). General-relativistic visualization. Computing in Science and Engineering, 13(6), 64-70.

The above example of the museum exhibit links to other visualization activities for public outreach: 
Miguel Aragon-Calvo and Mark SubbaRao~\cite{Aragon-Calvo2015} described a virtual flight through the universe that addresses and balances scientific and artistic goals. For example, they discussed how they adapted the virtual size of galaxies in order to address the issue of vast space between those in the visualization. Figure~\ref{fig:galaxies} shows a snapshot of a flight through the universe.

\begin{figure}[H]
    \begin{center}    
        \includegraphics[width=0.9\textwidth]{galaxies.png}
        \caption{Virtual flight through the universe. 
 Image: \copyright{} 2015 IEEE, reprinted from Aragon-Calvo and SubbaRao~\cite{Aragon-Calvo2015}.
  \label{fig:galaxies}}        
    \end{center}
\end{figure}

Another VizCorner paper looked into the issue of reconstructing 3D models of astronomical nebulae from 2D images along with the efficient volume rendering of such nebulae. Other papers discussed approaches to visualizing special relativity and general relativity.

\subsubsection{Further Applications and Visualization Techniques}
%Salinet, J., Oliveira, G., Vanheusden, F., Comba, J., Ng, G., & Schlindwein, F. (2013). Visualizing intracardiac atrial fibrillation electrograms using spectral analysis. Computing in Science and Engineering, 15(2), 79-87.
%Santos, E., Poco, J., Wei, Y., Liu, S., Cook, B., Williams, D., & Silva, C. (2013). UV-CDAT: Analyzing climate datasets from a user's perspective. Computing in Science and Engineering, 15(1), 94-103.
%Michalski, M., Rieth, M., Kempf, A., & Krüger, J. (1. 11 2017). CoFlaVis: A visualization system for pulverized coal flames. Computing in Science and Engineering, 19(6), 72-78.

The field of astronomy and astrophysics is only one relevant application area for visualization. The VizCorner acknowledges the breadth of possible applications. Other application examples covered in previous papers include the visualization of climate data sets, simulation of coal combustion processes in the form of pulverized coal flames, or the visualization of intracardiac atrial fibrillation electrograms. These examples show the full range of visualization applications from science to engineering to medical science. Also, they showcase the practical relevance of visualization---one of the goals of the VizCorner.

With multidimensional and multivariate visualization, flow visualization, and human-computer interaction, we have already touched some relevant visualization themes that build a versatile foundation for many applications. However, the VizCorner is not restricted to just these themes. Previous papers covered approaches to direct analytical support in visualization, advanced volume rendering, efficient parallel processing, mathematical modeling, or incorporating text in visualization.  %---showing the breadth in visualization methods relevant for science and engineering.

%Text:
%Koch, S., Heimerl, F., & Ertl, T. (2013). Visual document retrieval: Supporting text search and analysis with visual analytics. Computing in Science and Engineering, 15(4), 66-74.
%Latif, S., & Beck, F. (1. 11 2018). Visually augmenting documents with data. Computing in Science and Engineering, 20(6).

%Unclassified:
%Ament, M., Weiskopf, D., & Dachsbacher, C. (1. 3 2016). Ambient volume illumination. Computing in Science and Engineering, 18(2), 90-97.
%Sadlo, F., Üffinger, M., Pagot, C., Osmari, D., Comba, J., Ertl, T., . . . Weiskopf, D. (5 2011). Visualization of cell-based higher-order fields. Computing in Science and Engineering, 13(3), 84-91.
%Vo, H., Comba, J., Geveci, B., & Silva, C. (9 2011). Streaming-enabled parallel data flow framework in the visualization toolkit. Computing in Science and Engineering, 13(5), 72-82.
%Andrienko, G., Andrienko, N., Bosch, H., Ertl, T., Fuchs, G., Jankowski, P., & Thom, D. (2013). Thematic patterns in georeferenced tweets through space-time visual analytics. Computing in Science and Engineering, 15(3), 72-82.

\subsubsection{Summary and Reflections}

This very brief tour through past VizCorner articles shows the breadth of basic visualization techniques and the full range of exciting applications alike. We have seen some themes covered in more depth, such as flow visualization or multivariate data visualization---these are directly driven by typical visualization needs in science and engineering. However, the VizCorner also tries to bring in trends from the visualization research community that might have a rather long-term impact on science and engineering. For example, advanced human-computer interaction techniques might eventually improve how engineers or scientists work with their visual-interactive tools.

Two VizCorner articles~\cite{Scheidegger2011,Comba2017} have already provided an overview of trends in visualization and their impact on computing in science and engineering. They reported on the 2011 and 2016 IEEE VIS Conferences (wwww.ieeevis.org), which is the premier venue for visualization research. The two papers also included research trends that were present at that time. In particular, the 2016 IEEE VIS report~\cite{Comba2017} identified some trends that, in a similar form, are still relevant. Let us have a closer look at current trends in visualization research.

\section{Trends in Visualization}

We want to share our perspective on what we see as important, ongoing trends. We adopt the point of view from visualization research as a whole, and then relate this to the particular interests for computing in science and engineering.

\subsection{The Interplay of Artificial Intelligence Techniques and Visualization}

Artificial Intelligence (AI) techniques play a crucial role in recognition tasks. In particular, machine learning (ML) algorithms are very efficient in pattern recognition problems such as image classification, speech recognition, and sentiment analysis. Over the past years, we have seen an increasing intersection between ML and visualization with the popularity of Deep Learning (DL). Unlike classical ML algorithms, DL algorithms process sets of labeled samples using multiple layers to extract features automatically.
A common intersection between AI and visualization is related to the field of explainable AI, where we use visualization techniques to gain insights into the inner workings of DL neural networks. We have recently seen several works that describe techniques to interpret DL models~\cite{hohman2018visual,GARCIA2018}. Most approaches aim at understanding how features are perceived during the learning process, in which visualization techniques can offer insights to improve the process.

The complexity of DL architectures also benefits from visualization to identify which aspects of the network can be modified. Finally, the understanding of the inner working of a deep model, a process called training analysis, has also been drawing attention recently by including the human in the loop. Such techniques are being referred to as interactive visual machine learning (IVML). As the popularity of DL grows steadily, we foresee that visualization techniques will continue to play a crucial role in understanding and improving new approaches.

With the increasing use of ML and especially DL in science and engineering, this trend is highly relevant for us as well.

\subsection{Getting Insights from Data: Visualization in Data Science}

Data is at the center of essential tasks in our society, such as decision-making, planning strategies, or process optimizations, and of course, also in data-driven science and engineering. The extraction of insights from data is the main focus of data science. There are, however, challenges in this process, since data is often massive, heterogeneous, subject to errors, and comes in a streaming fashion. Along with other fields such as applied mathematics, statistics, ML (see above), and computational modeling, visualization plays a fundamental part in data analysis. Urban data is a good example that can benefit from visualization techniques, such as spatio-temporal representations of data~\cite{Doraiswamy2018}. Several works describe ways to understand mobility patterns in cities from their fleet of taxis, trains, or bike-sharing systems. 
%Sports is another source of data that significantly benefit from visualization techniques. 
%Several approaches use visualization to understand soccer, tennis, and baseball matches, among other sports~\cite{perin2018}. 
Another example is social media. Visualization techniques are being used to display and explore social networks and their flow of information diffusion, or even explore the textual content shared among users~\cite{Chen2017}. Visual text analysis, which combines visualization along with other areas such as data mining, natural language processing, and machine learning, offers great ways to extract insights from social media. In all these examples we see how visualization techniques allow the extraction of insights from large, heterogeneous and unstructured data. In science and engineering we observe a growing need of this type of data and similar needs for insights.

%{\bf TODO: Link to science and engineering?! Especially since the last few lines are not really on CiSE topics.}

\subsection{Integration of VIS Subfields}

IEEE VIS has been structured around its three main conferences: the IEEE Conference on Visual Analytics Science and Technology (VAST), the IEEE Conference on Information Visualization (InfoVis), and the IEEE Conference on Scientific Visualization (SciVis). These conferences have traditionally targeted different subfields of visualization. For example, scientific visualization primarily considers data from scientific applications, i.e., it is the natural link to computing in science and engineering. Scientific visualization usually deals with data with a given spatial embedding such as 3D scalar or vector fields. In contrast, information visualization typically works data without a given spatial mapping; for example, the visualization of networks or high-dimensional data. Melanie Tory and Torsten M{\"o}ller provide more details on spatial and nonspatial visualization \cite{ToryM04}. Finally, visual analytics often includes ML techniques for improved data analysis and addresses the challenges of data science.

We have already seen in the first part of this article that the separation of the three subfields is not so clear in practice. For example, multidimensional or multivariate data is highly interesting in science and engineering and, therefore, appropriate visualizations will likely include both SciVis and InfoVis techniques. We have seen a lot of highly exciting papers that combine techniques from the subfields of VIS; these are particularly interesting for computing in science and engineering. 
IEEE VIS reflects this trend toward combined techniques by integrating the VAST, InfoVis, and SciVis Conferences, according to the results of a restructuring process (http://ieeevis.org/governance/restructuring).

Besides this integration, we have also been witnessing visualization as a growing and maturing field. Therefore, fundamental and long-standing research challenges are still being addressed, but now on a more refined and advanced level. As an example that is particularly relevant for computing in science and engineering, there is ongoing research on making visualization useful for applications in high-performance computing (HPC), by exploiting in-situ visualization integrated into the HPC environment or developing efficient, parallel visualization techniques. Another ongoing topic targets improved mathematical methods for visualization, for example, relying on topological techniques or uncertainty quantification.

In summary, we expect that the integration of the subfields of visualization will continue and deliver improved visualization systems. We also see ongoing research in several of these subfields that are relevant for applications in science and engineering.

\subsection{Visualization and Human-Computer Interaction}

We have also been witnessing an increasingly strong link between visualization and human-computer interaction (HCI) research. One theme of our retrospective addresses HCI from the visualization perspective. With interactive data exploration, interaction techniques play a highly significant role in visualization, with a large number of papers at IEEE VIS. However, visualization has also become interesting for HCI research as a relevant topic. This is reflected by the fact that the leading HCI conference---ACM CHI (https://chi.acm.org)---included a separate subcommittee for visualization for CHI 2020. This link nicely strengthens the human-centered perspective on visualization.
Furthermore, it is the right way of addressing the long-standing and challenging issue of evaluating visualization approaches from the user's perspective. For example, the series of BELIV Workshops (https://beliv-workshop.github.io/) targets this issue. While these HCI and user-based evaluation problems are primarily interesting for basic visualization research, we expect that improved interactive visualization will eventually impact any visualization application, including ones in science and engineering.

\section{Conclusion}

We think that our discussion of past VizCorner papers from the last ten years and the outlook on trends show that visualization is a thriving and evolving research field with many interesting connections to computing in science and engineering. With this paper, the tenure of one of the current co-editors (Daniel Weiskopf) will end. 
I, Daniel, have been serving in this role since 2011, and it has been a great pleasure working for the VizCorner. Thanks to all readers and authors of the VizCorner, and special thanks to my former co-editor Claudio Silva, my current co-editor Joao Comba, and the whole CiSE team!

The VizCorner with continue with a new incoming co-editor joining Joao. This change of co-editor will allow us to refresh the topics of the VizCorner. We look forward to bringing interesting new articles in the years to come.

%TODO: Joao: anything on call for papers? Future plans?
%With visualization thriving as an important topic for computing in science ...

\bibliographystyle{IEEEtran}
\bibliography{refs}

\textbf{Joao Comba} is a full professor of computer science at UFRGS, Brazil. His research interests include visual analytics, data science, information visualization, geometric algorithms, spatial data structures and high-performance computing. Joao Comba received a Ph.D. in computer science from Stanford University, USA. Contact him at comba@inf.ufrgs.br.

\textbf{Daniel Weiskopf} is a professor of computer science at the University of Stuttgart, Germany. His research interests include visualization, visual analytics, eye tracking, computer graphics, and special and general relativity. Weiskopf received a Dr. rer. nat. (PhD) in physics from the University of T\"ubingen, Germany. Contact him at weiskopf@visus.uni-stuttgart.de.

\end{document}

### 2020-03-27 CombaWeiskopf2020/paper.tex
fix latex errors and wide figures
move intro blurb to abstract

@@
 \documentclass[10pt,journal,compsoc]{IEEEtran}
+
+\usepackage{cite}
+\usepackage[pdftex]{graphicx}
 
@@
 	\IEEEcompsocthanksitem Daniel Weiskopf is at University of Stuttgart, Germany.
-}
+}}
 
@@
 \begin{abstract}
-We provide a retrospective of the Visualization Corner and how its articles covered visualization topics since 2011.
-This leads us to identify current and future developments in visualization research and their impact on computing in science and engineering.
+A retrospective of the Visualization Corner (VizCorner) since 2011 leads us to identify current and future developments in visualization research and their impact on computing in science and engineering.
+The VizCorner department of CiSE reflects the long-term and fruitful overlap between computer-based visualization and computational science and engineering.
+As early as 1987, the Workshop on Visualization in Scientific Computing identified the need for visualization for the growing area of simulation-oriented science~\cite{NSFreport1987}.
+Visualization now plays a key role in visual data analysis and communication in science, engineering, and beyond.
+Looking into what we have covered in VizCorner for the last decade, we connect with current trends in visualization research, leading to possible future developments for visualization in science and engineering.
 \end{abstract}
@@
 
-The Visualization Corner (VizCorner) is a long-standing department in IEEE Computing in Science \& Engineering (CiSE).
-It reflects the long-term and fruitful overlap between computer-based visualization and computational science and engineering.
-As early as 1987, the Workshop on Visualization in Scientific Computing identified the need for visualization for the growing area of simulation-oriented science~\cite{NSFreport1987}.
-In general, visualization has been playing a great role in visual data analysis and communication in science, engineering, and beyond.
-
-Let us first focus on the scope of the VizCorner by looking into what we have covered there in the last decade.
-Then we will relate this to current trends in visualization research, leading to possible future developments for visualization in science and engineering.
-
 \section{Retrospective of the VizCorner}
 
-The VizCorner has three column articles per year on average.
-Our goal is to cover the breadth of visualization research and practice relevant to the readers of this magazine.
-Therefore, the VizCorner wants to provide a glimpse of basic visualization research questions but also illustrate and highlight the practical impact of visualization.
-We have been soliciting papers mostly from the visualization community in order to bring in their perspective to the various areas of computational science and engineering.
-
-Let us organize the papers from 2011 until now into themes, indicating the focal areas of the VizCorner. 
-We will provide examples of previous papers that illustrate the breadth of visualization possibilities.
+The goal of the VizCorner is to cover the breadth of visualization research and practice relevant to the readers of this magazine.
+We provide a glimpse of basic research questions and highlight the practical impact of visualization.
+Organizing the papers from 2011 until now into themes highlights the focal areas of the VizCorner, 
+and illustrates the breadth of visualization possibilities.
 
@@
 
-\begin{figure}[H]
+\begin{figure*}
     \begin{center}    
@@
     \end{center}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[th]
+\begin{figure*}
 	\begin{center}	
@@
 	\end{center}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[th]
+\begin{figure*}
     \begin{center}    
@@
 \bibliographystyle{IEEEtran}
-\bibliography{refs}
+\bibliography{./refs}
+
+\bigskip
 

### 2020-03-27 CombaWeiskopf2020/paper.tex
edits section 1

@@
 
-\subsubsection{Multidimensional and Multivariate Data Visualization}
+\subsection*{Multidimensional and multivariate data visualization}
 % 4 papers:
@@
 
-Multidimensional and multivariate data is almost omnipresent in computational science and engineering. Accordingly, we had several papers addressing this topic.
-The paper by Helmut Doleisch and Helwig Hauser~\cite{Doleisch2012} nicely shows how the combination of 2D and 3D visualization, flexible degree-of-interest definition, and interactive exploration leads to an effective visual analysis of complex simulation data.
-They discuss, for example, how multiple simulation attributes from computational fluid dynamics can be visually analyzed.
+Multidimensional and multivariate data is almost omnipresent in computational science and engineering. 
+Helmut Doleisch and Helwig Hauser~\cite{Doleisch2012} nicely showed how the combination of 2D and 3D visualization, flexible degree-of-interest definition, and interactive exploration lead to effective visual analysis of complex simulation data.
+They discussed how multiple simulation attributes from computational fluid dynamics can be visually analyzed.
 Figure~\ref{fig:simvis} shows a screenshot from their software system, illustrating how different visualization views are integrated in order to support a user-oriented exploration and analysis process.
@@
         \includegraphics[width=0.9\textwidth]{simvis.png}
-        \caption{Example of multivariate data visualization with the SimVis system~\cite{Doleisch2012}.
+        \caption{Example of multivariate data visualization with the SimVis system.
 This screenshot shows the windflow around a building, where turbulent parts are emphasized using a vortex detection method.
@@
 
-Other papers for this theme covered multidimensional projection techniques that reduce multidimensional data to 2D data for visualization purposes.
-One paper specifically included user-centered control into the projection techniques.
-Another paper followed a different approach, employing parallel coordinates to show multiple data dimensions simultaneously.
-Overall, the papers in this category come with quite generic visualizations that can be useful in many areas of science and engineering.
+The VizCorner has also covered projection techniques that reduce multidimensional data to 2D data for visualization purposes, including user-centered control of the projection techniques, and showing multiple data dimensions simultaneously via parallel coordinates.
+Papers in this category often include generic visualizations that can be useful in many areas of science and engineering.
 
-\subsubsection{Flow Visualization}
+\subsubsection*{Flow visualization}
 % 3+1 papers:
@@
 
-This topic area is a classic one in scientific visualization, considering input from flow simulation and experiments alike.
-We pick one of the VizCorner papers as an example: Grzegorz Karch and colleagues~\cite{Karch2012} described methods for high-quality computational flow visualization utilizing the metaphor of dye advection.
+A classic research area, flow visualization considers input from flow simulation and experiments alike.
+Our pick from the VizCorner is by Grzegorz Karch and colleagues~\cite{Karch2012}, who described methods for high-quality flow visualization utilizing the metaphor of dye advection.
 Experimental flow visualization has a long tradition of dye-based observations.
-With its computational counterpart, we can use the same idea now in a computer-based visualization, see Figure~\ref{fig:karman} as a typical example.
+Its computational counterpart uses the same idea now in a computer-based visualization; see an example in Figure~\ref{fig:karman}.
 
@@
 		\includegraphics[width=0.9\textwidth]{karman_weno3.png}
-		\caption{Computational dye advection~\cite{Karch2012}, applied to a 2D simulation of a von K{\'a}rm{\'a}n vortex street. Image: \copyright{} 2012 IEEE, reprinted from Karch et al.~\cite{Karch2012}.
+		\caption{Computational dye advection applied to a 2D simulation of a von K{\'a}rm{\'a}n vortex street. Image: \copyright{} 2012 IEEE, reprinted from Karch et al.~\cite{Karch2012}.
   \label{fig:karman}}		
@@
 
-Other papers described alternative visual representations of flow, for example, by using textures or streak and timelines.
-Another paper highlighted interaction techniques with haptics that let users virtually touch the data.
+The VizCorner has covered alternative visual representations of flow using textures or streak and timelines, 
+and interaction techniques with haptics that let users virtually touch the data.
 
-\subsubsection{Human-Computer Interaction and Storytelling}
+\subsubsection*{Human-computer interaction and storytelling}
 
@@
 
-The latter paper leads us to the next theme that focuses on the human-centered aspects of visualization.
-In general, interaction plays an essential role in visualization; for example, in order to support scientists in exploring their simulation data.
-Other VizCorner articles highlighted new developments in eye tracking for interactive visualization or the role of human visual perception.
-Finally, Franz Sauer and colleagues~\cite{Sauer2016} went all the way to discussing visualization as a means of storytelling that can address a wide range of audiences, from domain scientists to general scientists to the public.
-For the latter one, they showcased an example of an interactive museum exhibit that visualized plankton populations to museum visitors, see Figure~\ref{fig:museum}.
+Another theme in the VizCorner focuses on the human-centered aspects of visualization.
+Interaction plays an essential role in visualization, supporting scientists in exploring their simulation data.
+The VizCorner has highlighted new developments in eye tracking for interactive visualization or the role of human visual perception.
+Franz Sauer and colleagues~\cite{Sauer2016} went all the way to discussing visualization as a means of storytelling that can address a wide range of audiences, from domain scientists to general scientists to the public.
+They showcased an example of an interactive museum exhibit that visualized plankton populations for visitors; see Figure~\ref{fig:museum}.
 
@@
 
-\subsubsection{Astronomy and Astrophysics}
+\subsubsection{Astronomy and astrophysics}
 
@@
 
-The above example of the museum exhibit links to other visualization activities for public outreach: 
+The museum exhibit mentioned above links to other visualization activities for public outreach: 
 Miguel Aragon-Calvo and Mark SubbaRao~\cite{Aragon-Calvo2015} described a virtual flight through the universe that addresses and balances scientific and artistic goals.
-For example, they discussed how they adapted the virtual size of galaxies in order to address the issue of vast space between those in the visualization.
+They discussed how they adapted the virtual size of galaxies in order to address the issue of vast space between them in the visualization.
 Figure~\ref{fig:galaxies} shows a snapshot of a flight through the universe.
@@
 
-Another VizCorner paper looked into the issue of reconstructing 3D models of astronomical nebulae from 2D images along with the efficient volume rendering of such nebulae.
+The VizCorner also looked into the issue of reconstructing 3D models of astronomical nebulae from 2D images along with the efficient volume rendering of such nebulae.
 Other papers discussed approaches to visualizing special relativity and general relativity.
 
-\subsubsection{Further Applications and Visualization Techniques}
+\subsubsection{Other applications and visualization techniques}
 %Salinet, J., Oliveira, G., Vanheusden, F., Comba, J., Ng, G., & Schlindwein, F.
@@
 
-The field of astronomy and astrophysics is only one relevant application area for visualization.
-The VizCorner acknowledges the breadth of possible applications.
-Other application examples covered in previous papers include the visualization of climate data sets, simulation of coal combustion processes in the form of pulverized coal flames, or the visualization of intracardiac atrial fibrillation electrograms.
-These examples show the full range of visualization applications from science to engineering to medical science.
-Also, they showcase the practical relevance of visualization---one of the goals of the VizCorner.
+The VizCorner encompasses the breadth of possible applications of visualization.
+We have covered the visualization of climate data sets, simulation of coal combustion processes in the form of pulverized coal flames, or the visualization of intracardiac atrial fibrillation electrograms.
+Showing the full range of visualization applications from science to engineering to medical science, and the practical relevance of visualization, is among the goals of the VizCorner.
 
 With multidimensional and multivariate visualization, flow visualization, and human-computer interaction, we have already touched some relevant visualization themes that build a versatile foundation for many applications.
-However, the VizCorner is not restricted to just these themes.
-Previous papers covered approaches to direct analytical support in visualization, advanced volume rendering, efficient parallel processing, mathematical modeling, or incorporating text in visualization.
+The VizCorner is not restricted to these themes, and we have covered approaches to direct analytical support in visualization, advanced volume rendering, efficient parallel processing, mathematical modeling, or incorporating text in visualization.
 %---showing the breadth in visualization methods relevant for science and engineering.
@@
 
-\subsubsection{Summary and Reflections}
+\subsubsection{Summary and reflections}
 
-This very brief tour through past VizCorner articles shows the breadth of basic visualization techniques and the full range of exciting applications alike.
-We have seen some themes covered in more depth, such as flow visualization or multivariate data visualization---these are directly driven by typical visualization needs in science and engineering.
-However, the VizCorner also tries to bring in trends from the visualization research community that might have a rather long-term impact on science and engineering.
+This brief tour through past VizCorner articles shows the breadth of basic visualization techniques and the full range of exciting applications.
+We have seen some themes covered in more depth, such as flow visualization or multivariate data visualization---these are driven by typical visualization needs in science and engineering.
+The VizCorner also tries to bring in trends from the visualization research community that might have a more long-term impact on science and engineering.
 For example, advanced human-computer interaction techniques might eventually improve how engineers or scientists work with their visual-interactive tools.
 
-Two VizCorner articles~\cite{Scheidegger2011,Comba2017} have already provided an overview of trends in visualization and their impact on computing in science and engineering.
-They reported on the 2011 and 2016 IEEE VIS Conferences (wwww.ieeevis.org), which is the premier venue for visualization research.
-The two papers also included research trends that were present at that time.
-In particular, the 2016 IEEE VIS report~\cite{Comba2017} identified some trends that, in a similar form, are still relevant.
+The VizCorner provided overviews of trends in visualization and the impact on computing in science and engineering, with synopsis of the 2011 and 2016 IEEE VIS Conferences (wwww.ieeevis.org), the premier venue for visualization research~\cite{Scheidegger2011,Comba2017}.
+The 2016 IEEE VIS report identified some trends that are still relevant.
 Let us have a closer look at current trends in visualization research.
 
-\section{Trends in Visualization}
+\section{Trends in visualization}
 

### 2020-03-27 CombaWeiskopf2020/paper.tex
edits section 2 and fin

@@
     \begin{center}    
-        \includegraphics[width=0.55\textwidth]{museum1.png} \hspace{-3em} \raisebox{12ex}{\includegraphics[width=0.5\textwidth]{museum2.png}}
+        \includegraphics[width=0.40\textwidth]{museum1.png} %\hspace{-3em} \raisebox{12ex}{\includegraphics[width=0.5\textwidth]{museum2.png}}
+        \includegraphics[width=0.48\textwidth]{museum2.png}
         \caption{Interactive table with a plankton visualization.
@@
 We want to share our perspective on what we see as important, ongoing trends.
-We adopt the point of view from visualization research as a whole, and then relate this to the particular interests for computing in science and engineering.
+We adopt the point of view of visualization research as a whole, and address the particular interests for computing in science and engineering.
 
-\subsection{The Interplay of Artificial Intelligence Techniques and Visualization}
+\subsection*{Interplay of artificial intelligence and visualization}
 
 Artificial Intelligence (AI) techniques play a crucial role in recognition tasks.
-In particular, machine learning (ML) algorithms are very efficient in pattern recognition problems such as image classification, speech recognition, and sentiment analysis.
+In particular, machine learning (ML) algorithms are efficient in pattern-recognition problems such as image classification, speech recognition, and sentiment analysis.
 Over the past years, we have seen an increasing intersection between ML and visualization with the popularity of Deep Learning (DL).
@@
 The complexity of DL architectures also benefits from visualization to identify which aspects of the network can be modified.
-Finally, the understanding of the inner working of a deep model, a process called training analysis, has also been drawing attention recently by including the human in the loop.
+Finally, the understanding of the inner working of a deep model, a process called training analysis, has also been drawing attention recently by including the human-in-the-loop.
 Such techniques are being referred to as interactive visual machine learning (IVML).
@@
 
-With the increasing use of ML and especially DL in science and engineering, this trend is highly relevant for us as well.
-
-\subsection{Getting Insights from Data: Visualization in Data Science}
+\subsection*{Getting insights from data: visualization in data science}
 
 Data is at the center of essential tasks in our society, such as decision-making, planning strategies, or process optimizations, and of course, also in data-driven science and engineering. 
-The extraction of insights from data is the main focus of data science.
-There are, however, challenges in this process, since data is often massive, heterogeneous, subject to errors, and comes in a streaming fashion.
+Extracting insights from data is challenging because data is often massive, heterogeneous, subject to errors, and comes in a streaming fashion.
 Along with other fields such as applied mathematics, statistics, ML (see above), and computational modeling, visualization plays a fundamental part in data analysis.
@@
 In all these examples we see how visualization techniques allow the extraction of insights from large, heterogeneous and unstructured data.
-In science and engineering we observe a growing need of this type of data and similar needs for insights.
+In science and engineering we observe a growing need for this type of data and similar needs for insights.
 
@@
 
-\subsection{Integration of VIS Subfields}
+\subsection*{Integration of VIS subfields}
 
 IEEE VIS has been structured around its three main conferences: the IEEE Conference on Visual Analytics Science and Technology (VAST), the IEEE Conference on Information Visualization (InfoVis), and the IEEE Conference on Scientific Visualization (SciVis).
-These conferences have traditionally targeted different subfields of visualization.
-For example, scientific visualization primarily considers data from scientific applications, i.e., it is the natural link to computing in science and engineering.
-Scientific visualization usually deals with data with a given spatial embedding such as 3D scalar or vector fields.
-In contrast, information visualization typically works data without a given spatial mapping; for example, the visualization of networks or high-dimensional data.
-Melanie Tory and Torsten M{\"o}ller provide more details on spatial and nonspatial visualization \cite{ToryM04}.
+Scientific visualization primarily considers data from scientific applications, i.e., it is the natural link to computing in science and engineering.
+It usually deals with data with a given spatial embedding such as 3D scalar or vector fields.
+Information visualization typically works with data lacking a spatial mapping; for example, the visualization of networks or high-dimensional data.
+Melanie Tory and Torsten M{\"o}ller provide details on spatial and non-spatial visualization \cite{ToryM04}.
 Finally, visual analytics often includes ML techniques for improved data analysis and addresses the challenges of data science.
 
-We have already seen in the first part of this article that the separation of the three subfields is not so clear in practice.
+The separation of the three subfields is not so clear in practice.
 For example, multidimensional or multivariate data is highly interesting in science and engineering and, therefore, appropriate visualizations will likely include both SciVis and InfoVis techniques.
-We have seen a lot of highly exciting papers that combine techniques from the subfields of VIS; these are particularly interesting for computing in science and engineering.
+We have seen a lot of exciting papers that combine techniques from the subfields of VIS; these are particularly interesting for computing in science and engineering.
 IEEE VIS reflects this trend toward combined techniques by integrating the VAST, InfoVis, and SciVis Conferences, according to the results of a restructuring process (http://ieeevis.org/governance/restructuring).
@@
 Besides this integration, we have also been witnessing visualization as a growing and maturing field.
-Therefore, fundamental and long-standing research challenges are still being addressed, but now on a more refined and advanced level.
+Fundamental and long-standing research challenges are still being addressed, but now on a more refined and advanced level.
 As an example that is particularly relevant for computing in science and engineering, there is ongoing research on making visualization useful for applications in high-performance computing (HPC), by exploiting in-situ visualization integrated into the HPC environment or developing efficient, parallel visualization techniques.
@@
 
-\subsection{Visualization and Human-Computer Interaction}
+\subsection*{Visualization and human-computer interaction}
 
@@
 One theme of our retrospective addresses HCI from the visualization perspective.
-With interactive data exploration, interaction techniques play a highly significant role in visualization, with a large number of papers at IEEE VIS.
-However, visualization has also become interesting for HCI research as a relevant topic.
-This is reflected by the fact that the leading HCI conference---ACM CHI (https://chi.acm.org)---included a separate subcommittee for visualization for CHI 2020.
-This link nicely strengthens the human-centered perspective on visualization.
-Furthermore, it is the right way of addressing the long-standing and challenging issue of evaluating visualization approaches from the user's perspective.
+With interactive data exploration, interaction techniques play an important role in visualization, with a large number of papers at IEEE VIS.
+Visualization has also become interesting for HCI research.
+This is reflected by the fact that the leading HCI conference---ACM CHI (https://chi.acm.org)---included in 2020 a separate subcommittee for visualization.
+This link strengthens the human-centered perspective on visualization.
+It is the right way of addressing the long-standing and challenging issue of evaluating visualization approaches from the user's perspective.
 For example, the series of BELIV Workshops (https://beliv-workshop.github.io/) targets this issue.
@@
 
-We think that our discussion of past VizCorner papers from the last ten years and the outlook on trends show that visualization is a thriving and evolving research field with many interesting connections to computing in science and engineering.
+Our retrospective of the VizCorner in the last ten years and the outlook on trends show that visualization is a thriving and evolving research field with many interesting connections to computing in science and engineering.
 With this paper, the tenure of one of the current co-editors (Daniel Weiskopf) will end.
@@
 
-The VizCorner with continue with a new incoming co-editor joining Joao.
-This change of co-editor will allow us to refresh the topics of the VizCorner.
+The VizCorner will continue with a new incoming co-editor joining Joao, who will help us to refresh the topics covered in the department.
 We look forward to bringing interesting new articles in the years to come.

### 2020-03-28 CombaWeiskopf2020/paper.tex
delete TODO note
remove subsection numbers
reorg: no refs in abstract

@@
 \begin{abstract}
-A retrospective of the Visualization Corner (VizCorner) since 2011 leads us to identify current and future developments in visualization research and their impact on computing in science and engineering.
-The VizCorner department of CiSE reflects the long-term and fruitful overlap between computer-based visualization and computational science and engineering.
-As early as 1987, the Workshop on Visualization in Scientific Computing identified the need for visualization for the growing area of simulation-oriented science~\cite{NSFreport1987}.
-Visualization now plays a key role in visual data analysis and communication in science, engineering, and beyond.
+A retrospective of the CiSE Visualization Corner (VizCorner) department since 2011 leads us to identify current and future developments in visualization research and their impact on computing in science and engineering.
+The VizCorner  reflects the long-term and fruitful overlap between computer-based visualization and computational science and engineering. 
 Looking into what we have covered in VizCorner for the last decade, we connect with current trends in visualization research, leading to possible future developments for visualization in science and engineering.
@@
 
-The goal of the VizCorner is to cover the breadth of visualization research and practice relevant to the readers of this magazine.
+The CiSE Visualization Corner (VizCorner) department covers the breadth of visualization research and practice relevant to the readers of this magazine.
 We provide a glimpse of basic research questions and highlight the practical impact of visualization.
+As early as 1987, the Workshop on Visualization in Scientific Computing identified the need for visualization for the growing area of simulation-oriented science~\cite{NSFreport1987}.
+Visualization now plays a key role in visual data analysis and communication in science, engineering, and beyond.
 Organizing the papers from 2011 until now into themes highlights the focal areas of the VizCorner, 
@@
         \includegraphics[width=0.48\textwidth]{museum2.png}
-        \caption{Interactive table with a plankton visualization.
-Image: \copyright{} 2016 IEEE, reprinted from Sauer et al.~\cite{Sauer2016}.
-TODO for layout: could you place the right image slightly overlapping on the left on?
-As to imply something similar to an inset?
+        \caption{Interactive table with a plankton visualization. Image: \copyright{} 2016 IEEE, reprinted from Sauer et al.~\cite{Sauer2016}. 
   \label{fig:museum}}        
@@
 
-\subsubsection{Astronomy and astrophysics}
+\subsubsection*{Astronomy and astrophysics}
 
@@
 
-\subsubsection{Other applications and visualization techniques}
+\subsubsection*{Other applications and visualization techniques}
 %Salinet, J., Oliveira, G., Vanheusden, F., Comba, J., Ng, G., & Schlindwein, F.
@@
 
-\subsubsection{Summary and reflections}
+\subsubsection*{Summary and reflections}
 

### 2020-03-28 CombaWeiskopf2020/paper.tex
remove redundancy

@@
 \begin{abstract}
-A retrospective of the CiSE Visualization Corner (VizCorner) department since 2011 leads us to identify current and future developments in visualization research and their impact on computing in science and engineering.
-The VizCorner  reflects the long-term and fruitful overlap between computer-based visualization and computational science and engineering. 
-Looking into what we have covered in VizCorner for the last decade, we connect with current trends in visualization research, leading to possible future developments for visualization in science and engineering.
+The CiSE Visualization Corner (VizCorner) department  reflects the long-term and fruitful overlap between computer-based visualization and computational science and engineering. 
+Looking into what the VizCorner has covered in the last decade, we identify current trends in visualization research,  and possible future developments for visualization in science and engineering.
 \end{abstract}
@@
 The CiSE Visualization Corner (VizCorner) department covers the breadth of visualization research and practice relevant to the readers of this magazine. 
-We provide a glimpse of basic research questions and highlight the practical impact of visualization.
 As early as 1987, the Workshop on Visualization in Scientific Computing identified the need for visualization for the growing area of simulation-oriented science~\cite{NSFreport1987}.

### 2020-04-05 CombaWeiskopf2020/paper.tex
edit fig.3 includegraphics
edits sent by Daniel

@@
 
-The VizCorner has also covered projection techniques that reduce multidimensional data to 2D data for visualization purposes, including user-centered control of the projection techniques, and showing multiple data dimensions simultaneously via parallel coordinates.
+The VizCorner has also covered projection techniques that reduce multidimensional data to 2D data for visualization purposes or methods that show multiple data dimensions simultaneously via parallel coordinates.
 Papers in this category often include generic visualizations that can be useful in many areas of science and engineering.
@@
 
-A classic research area, flow visualization considers input from flow simulation and experiments alike.
-Our pick from the VizCorner is by Grzegorz Karch and colleagues~\cite{Karch2012}, who described methods for high-quality flow visualization utilizing the metaphor of dye advection.
+A classic research area of scientific visualization, flow visualization considers input from flow simulation and experiments alike.
+Our pick from the VizCorner is by Grzegorz Karch and colleagues~\cite{Karch2012}, who described methods for high-quality computational flow visualization utilizing the metaphor of dye advection.
 Experimental flow visualization has a long tradition of dye-based observations.
@@
     \begin{center}    
-        \includegraphics[width=0.40\textwidth]{museum1.png} %\hspace{-3em} \raisebox{12ex}{\includegraphics[width=0.5\textwidth]{museum2.png}}
-        \includegraphics[width=0.48\textwidth]{museum2.png}
+        \includegraphics[width=\textwidth]{museum.png} 
         \caption{Interactive table with a plankton visualization. Image: \copyright{} 2016 IEEE, reprinted from Sauer et al.~\cite{Sauer2016}. 

### 2020-04-05 CombaWeiskopf2020/paper.tex
remove problematic sentence
delete sentence
final edits sent by Daniel

@@
 
-Another theme in the VizCorner focuses on the human-centered aspects of visualization.
 Interaction plays an essential role in visualization, supporting scientists in exploring their simulation data.
@@
 The VizCorner provided overviews of trends in visualization and the impact on computing in science and engineering, with synopsis of the 2011 and 2016 IEEE VIS Conferences (wwww.ieeevis.org), the premier venue for visualization research~\cite{Scheidegger2011,Comba2017}.
-The 2016 IEEE VIS report identified some trends that are still relevant.
 Let us have a closer look at current trends in visualization research.
@@
 The complexity of DL architectures also benefits from visualization to identify which aspects of the network can be modified.
-Finally, the understanding of the inner working of a deep model, a process called training analysis, has also been drawing attention recently by including the human-in-the-loop.
+Finally, the understanding of the inner working of a deep model, a process called training analysis, has also been drawing attention recently by including the human in the loop.
 Such techniques are being referred to as interactive visual machine learning (IVML).
@@
 Visual text analysis, which combines visualization along with other areas such as data mining, natural language processing, and machine learning, offers great ways to extract insights from social media.
-In all these examples we see how visualization techniques allow the extraction of insights from large, heterogeneous and unstructured data.
-In science and engineering we observe a growing need for this type of data and similar needs for insights.
+In all these examples, we see how visualization techniques allow the extraction of insights from large, heterogeneous, and unstructured data.
+In science and engineering, we observe a growing need for this type of data and similar needs for insights.
 
@@
 It usually deals with data with a given spatial embedding such as 3D scalar or vector fields.
-Information visualization typically works with data lacking a spatial mapping; for example, the visualization of networks or high-dimensional data.
+Information visualization typically works with data lacking a given spatial mapping; for example, the visualization of networks or high-dimensional data.
 Melanie Tory and Torsten M{\"o}ller provide details on spatial and non-spatial visualization \cite{ToryM04}.
@@
 Visualization has also become interesting for HCI research.
-This is reflected by the fact that the leading HCI conference---ACM CHI (https://chi.acm.org)---included in 2020 a separate subcommittee for visualization.
+This is reflected by the fact that the leading HCI conference---ACM CHI (https://chi.acm.org)---now includes a separate subcommittee for visualization.
 This link strengthens the human-centered perspective on visualization.

### 2020-06-25 Thyng2020/paper.tex
initial commit

\documentclass[10pt,journal,compsoc]{IEEEtran}

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{natbib}
\usepackage{url}
\usepackage{hyperref}
% \usepackage{breakurl}

\begin{document}

\title{The importance of colormaps}

\author{%
	Kristen M. Thyng
	\IEEEcompsocitemizethanks{
	\IEEEcompsocthanksitem Kristen Thyng is at Texas A\&M University in Oceanography.\protect
}}

\IEEEtitleabstractindextext{%
\begin{abstract}
Colormaps control the way data is understood when plotted because they map data values to color. When the colors used do not match the way humans perceive color, there can be a mismatch between interpretation and the data. Here I present relevant background information as well as provide suggestions as to improving colormap choices.
\end{abstract}
}

% make the title area
\maketitle

\section{Introduction}

The colors chosen to represent data in a colormap have a large influence on accurate data interpretation (Fig.~\ref{fig:motivation}). Given this, it is worth careful consideration of the colormap chosen for this representation, beyond the default available or what is familiar.

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/illustrator/draft3/motivation.pdf}
	\caption{}
	\label{fig:motivation}
\end{figure*}

\nocite{treinish1998should}
\nocite{borkin2011evaluation}
\nocite{thyng2016true}

Color is represented by three dimensions, called a colorspace. All colorspaces approximate how humans see color and use different dimensions for this representation. If a colormap is created from a perceptually-uniform colorspace, it has the special property that any equal distance along the colormap is perceived by humans as approximately equal in magnitude of change (Fig.~\ref{fig:evaluation}). A consequence of not using a perceptually-uniform colormap is perceptual jumps across the colormap, which can be large. These perceptual jumps introduce fake gradients into the data representation as well as obscure detail. Another important finding from previous studies is that humans best understand the relative form of data when it is mapped by lightness rather than hue \citep{kovesi2015good}.

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/illustrator/draft3/evaluation.pdf}
	\caption{}
	\label{fig:evaluation}
\end{figure*}
\nocite{kovesi2015good}
\nocite{viscm}

Colormaps for plotting relative data (rather than categorical data) can be generally grouped into 3 categories. First, sequential colormaps have monotonic lightness change and should be used to represent data that increases in size (such as chlorophyll, Fig.~\ref{fig:howtochoose}). Second, diverging colormaps have monotonic increase or decrease in lightness to a point, then change to monotonically decrease or increase. The utility of a diverging colormap is in representing data values relative to a critical point (for example, sea surface height relative to mean sea level). Third, cyclic colormaps should be used to represent data that cycles around a circle, like tidal phase. They should either not vary in lightness at all and cycle through hue only, or should cycle in both lightness and hue such that there is no major perceptual break in the colormap.

There are many considerations that should be taken into account when choosing the appropriate colormap for data presentation. For example, some form of color-blindness impacts about 7\% of Northern European, 4\% of Asian, and 3\% of African-descendent men \citep{sharpe1999opsin}. Avoiding colormaps that include both red and green, which look similar with the most common form of colorblindness, would be wise to invite a wider audience to the plot.

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/illustrator/draft3/howtochoose.pdf}
	\caption{}
	\label{fig:howtochoose}
\end{figure*}
% Data sources
\nocite{chlor}
\nocite{ssh}
\nocite{egbert2002efficient}

A wide variety of reasonable, available colormaps are presented in Fig.~\ref{fig:allmaps}. Colormaps are plotted by the lightness as calculated in perceptually-uniform colorspace \textit{CAM02-UCS} in order to demonstrate this important property for making a colormap choice.

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/illustrator/draft3/allmaps.pdf}
	\caption{}
	\label{fig:allmaps}
\end{figure*}
% colormaps
\nocite{matplotlib}
\nocite{nunez2018optimizing}
\nocite{cubehelix}
\nocite{twilight}
\nocite{matlab}
\nocite{mycarta}
\nocite{light2004end}
\nocite{brewer}
\nocite{crameri2018geodynamic}

\section{Conclusion}

Colormaps have fundamental control over interpretation of data, and thus should be given some attention. Background information based on human perception of color as well as some basic design advice has been presented to guide selections.

\bibliographystyle{IEEEtran}
\bibliography{./refs}

\bigskip

\textbf{Kristen Thyng} is an assistant research professor of oceanography at Texas A\&M University. Her research interests include coastal physics, numerical modeling, visualization, and data science. She received a Ph.D. in mechanical engineering from the University of Washington, USA. Contact her at kthyng@tamu.edu.

\end{document}

### 2020-06-25 Thyng2020/paper.tex
updates to text and small change to motivation fig

@@
 
-\usepackage{cite}
+% \usepackage{cite}
 \usepackage[pdftex]{graphicx}
 \usepackage{natbib}
-\usepackage{url}
+% \usepackage{url}
 \usepackage{hyperref}
 % \usepackage{breakurl}
+
+% \newcommand{\cm}[1]{\textbf{#1}}
 
@@
 
-The colors chosen to represent data in a colormap have a large influence on accurate data interpretation (Fig.~\ref{fig:motivation}).
-Given this, it is worth careful consideration of the colormap chosen for this representation, beyond the default available or what is familiar.
+The colors chosen to represent data in a colormap have a large influence on accurate data interpretation.
+Importantly, one study found an increase in mistakes in medical scan interpretation by domain experts accustomed to the rainbow-based \textbf{jet} colormap when the scan data were displayed with \textbf{jet} as compared with another colormap \citep{Spence:1999ea}.
+Other studies have also shown an increase in interpretation mistakes when data is mapped with rainbow-based colormaps  \citep{borkin2011evaluation,Bryant:2014bh}.
+Given this, it is worth careful consideration of the colormap chosen for this representation, beyond the default available or what is familiar (Fig.~\ref{fig:motivation}).
+Note that the colormap information presented in this paper is regarding relative data (as opposed to categorical data) for which one want to understand the form and see detailed relationships in size.
+
+Given the difficulty with rainbow-based colormaps like \textbf{jet}, one might consider using a grayscale map for everything.
+While a well-made grayscale colormap can indeed be an excellent choice for data presentation, it is not the best colormap for all data.
+For example, grayscale would not be the best choice for representing topography and bathymetry in a single map, where introducing suggestive colors for land and water, respectively, would greatly add to the map.
+Using a distinct colormap for each variable allows for tailoring of the colormap to unique properties of each variable as well as acclimatizing viewers to the custom colormap used per variable (more on this in Fig.~\ref{fig:howtochoose}).
 
 \begin{figure*}
-	\includegraphics[width=\textwidth]{figures/illustrator/draft3/motivation.pdf}
+	\includegraphics[width=\textwidth]{figures/motivation.pdf}
 	\caption{}
@@
 \nocite{thyng2016true}
+\nocite{Spence:1999ea}
 
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{figures/illustrator/draft3/evaluation.pdf}
+	\includegraphics[width=\textwidth]{figures/evaluation.pdf}
 	\caption{}
@@
 
-Colormaps for plotting relative data (rather than categorical data) can be generally grouped into 3 categories.
-First, sequential colormaps have monotonic lightness change and should be used to represent data that increases in size (such as chlorophyll, Fig.~\ref{fig:howtochoose}).
-Second, diverging colormaps have monotonic increase or decrease in lightness to a point, then change to monotonically decrease or increase.
+There are many considerations that should be taken into account when choosing the appropriate colormap for data presentation (Fig.~\ref{fig:howtochoose}):
+\begin{itemize}
+	\item \textbf{Use appropriate colormap type:} Colormaps for plotting relative data can be generally grouped into 3 categories.
+First, sequential colormaps have monotonic lightness change and should be used to represent data that increases in size (such as chlorophyll, see example in Fig.~\ref{fig:howtochoose}).
+Second, diverging colormaps have a monotonic increase or decrease in lightness to a point, then change to monotonically decrease or increase.
 The utility of a diverging colormap is in representing data values relative to a critical point (for example, sea surface height relative to mean sea level).
@@
 They should either not vary in lightness at all and cycle through hue only, or should cycle in both lightness and hue such that there is no major perceptual break in the colormap.
-
-There are many considerations that should be taken into account when choosing the appropriate colormap for data presentation.
-For example, some form of color-blindness impacts about 7\% of Northern European, 4\% of Asian, and 3\% of African-descendent men \citep{sharpe1999opsin}.
+	\item \textbf{Perceptual uniformity:} Use colormaps with perceptual uniformity in mind when possible, for example, made from a perceptually-uniform colorspace like \textit{CAM02-UCS}.
+	\item \textbf{Use intuition:} Use shades of green to represent changes in chlorophyll, or a colormap that goes from cool to warm colors to represent temperature.
+	\item \textbf{Color blindness:} Some form of color-blindness impacts about 7\% of Northern European, 4\% of Asian, and 3\% of African-descendent men \citep{sharpe1999opsin}.
 Avoiding colormaps that include both red and green, which look similar with the most common form of colorblindness, would be wise to invite a wider audience to the plot.
+	\item \textbf{Match colormap to data:} Choose one colormap per variable, considering the items in this list, and then stick to it for the whole presentation or paper.
+This way the audience will become accustomed to the meaning of the colors and be able to quickly determine meaning from plots without extensive study of the colorbar and legend.
+	\item \textbf{Magnitude vs.
+range:} Some sequential variables have a clear meaning of magnitude, like turbidity or turbulence, and others measure a metric, like temperature.
+For the former, consider a colormap with shades of one color, where increasing shades represent increasing size of the variable.
+For the latter, consider a colormap that transverses several colors so that there is always a color representing the range of data values instead of white, which could connote the absense of the variable.
+\end{itemize}
 
 \begin{figure*}
-	\includegraphics[width=\textwidth]{figures/illustrator/draft3/howtochoose.pdf}
+	\includegraphics[width=\textwidth]{figures/howtochoose.pdf}
 	\caption{}
@@
 \begin{figure*}
-	\includegraphics[width=\textwidth]{figures/illustrator/draft3/allmaps.pdf}
+	\includegraphics[width=\textwidth]{figures/allmaps.pdf}
 	\caption{}
@@
 % colormaps
-\nocite{matplotlib}
+\nocite{hunter2007matplotlib}
 \nocite{nunez2018optimizing}
@@
 
+\section{Acknowledgments}
+
+The notebook used to create the base figures in this paper can be found at \url{https://github.com/kthyng/cise-colormaps}.
+
+Plots were created using the following Python packages: matplotlib \citep{hunter2007matplotlib}, Jupyter \citep{kluyver2016jupyter}, numpy \citep{oliphant2006guide,walt2011numpy},
+xarray \citep{hoyer2017xarray},
+pandas \citep{mckinney-proc-scipy-2010}, cartopy \citep{Cartopy}, and colorspacious \citep{colorspacious} along with the colormaps packages cmocean, colorcet \citep{colorcet}, and palettable \citep{palettable}.
+
 \bibliographystyle{IEEEtran}

### 2020-06-26 Thyng2020/paper.tex
updates after feedback

@@
 The colors chosen to represent data in a colormap have a large influence on accurate data interpretation.
-Importantly, one study found an increase in mistakes in medical scan interpretation by domain experts accustomed to the rainbow-based \textbf{jet} colormap when the scan data were displayed with \textbf{jet} as compared with another colormap \citep{Spence:1999ea}.
-Other studies have also shown an increase in interpretation mistakes when data is mapped with rainbow-based colormaps  \citep{borkin2011evaluation,Bryant:2014bh}.
-Given this, it is worth careful consideration of the colormap chosen for this representation, beyond the default available or what is familiar (Fig.~\ref{fig:motivation}).
-Note that the colormap information presented in this paper is regarding relative data (as opposed to categorical data) for which one want to understand the form and see detailed relationships in size.
+One study found more mistakes in medical scan interpretation by domain experts accustomed to the rainbow-based \textbf{jet} colormap when the scan data were displayed with \textbf{jet} as compared with another colormap \citep{Spence:1999ea}.
+Other studies have also shown more interpretation mistakes when data is mapped with rainbow-based colormaps  \citep{borkin2011evaluation,Bryant:2014bh}.
+Given this, it is worth carefully considering the colormap chosen for data representation, beyond the default available or what is familiar (Fig.~\ref{fig:motivation}).
+The colormap information presented in this paper targets relative data (as opposed to categorical data) for which one want to understand the form and see detailed relationships in size.
 
-Given the difficulty with rainbow-based colormaps like \textbf{jet}, one might consider using a grayscale map for everything.
+Given the drawbacks of rainbow-based colormaps like \textbf{jet}, one might consider using a grayscale map for everything.
 While a well-made grayscale colormap can indeed be an excellent choice for data presentation, it is not the best colormap for all data.
-For example, grayscale would not be the best choice for representing topography and bathymetry in a single map, where introducing suggestive colors for land and water, respectively, would greatly add to the map.
-Using a distinct colormap for each variable allows for tailoring of the colormap to unique properties of each variable as well as acclimatizing viewers to the custom colormap used per variable (more on this in Fig.~\ref{fig:howtochoose}).
-
-\begin{figure*}
-	\includegraphics[width=\textwidth]{figures/motivation.pdf}
-	\caption{}
-	\label{fig:motivation}
-\end{figure*}
+For example, grayscale would not be the best choice for representing topography and bathymetry in a single map, where introducing suggestive colors for land and water, respectively, would greatly add to the interpretability of the map.
+Using a distinct colormap for each variable allows for tailoring of the colormap to unique properties of each variable as well as acclimatizing viewers to the custom colormap used per variable.
 
@@
 
-\begin{figure*}
-	\includegraphics[width=\textwidth]{figures/evaluation.pdf}
-	\caption{}
-	\label{fig:evaluation}
-\end{figure*}
 \nocite{kovesi2015good}
-\nocite{viscm}
 
-There are many considerations that should be taken into account when choosing the appropriate colormap for data presentation (Fig.~\ref{fig:howtochoose}):
+Many considerations should be taken into account when choosing the appropriate colormap for data presentation (Fig.~\ref{fig:howtochoose}):
 \begin{itemize}
@@
 They should either not vary in lightness at all and cycle through hue only, or should cycle in both lightness and hue such that there is no major perceptual break in the colormap.
-	\item \textbf{Perceptual uniformity:} Use colormaps with perceptual uniformity in mind when possible, for example, made from a perceptually-uniform colorspace like \textit{CAM02-UCS}.
+	\item \textbf{Perceptual uniformity:} Use colormaps with perceptual uniformity when possible, for example, made from a perceptually-uniform colorspace like \textit{CAM02-UCS}.
 	\item \textbf{Use intuition:} Use shades of green to represent changes in chlorophyll, or a colormap that goes from cool to warm colors to represent temperature.
@@
 For the former, consider a colormap with shades of one color, where increasing shades represent increasing size of the variable.
-For the latter, consider a colormap that transverses several colors so that there is always a color representing the range of data values instead of white, which could connote the absense of the variable.
+For the latter, consider a colormap that traverses several colors so that there is always a color representing the range of data values instead of white, which could connote the absense of the variable.
 \end{itemize}
 
-\begin{figure*}
-	\includegraphics[width=\textwidth]{figures/howtochoose.pdf}
-	\caption{}
-	\label{fig:howtochoose}
-\end{figure*}
-% Data sources
-\nocite{chlor}
-\nocite{ssh}
-\nocite{egbert2002efficient}
-
-A wide variety of reasonable, available colormaps are presented in Fig.~\ref{fig:allmaps}.
+A wide variety of reasonable, available colormaps are presented in Fig.~\ref{fig:allmaps} \citep{hunter2007matplotlib,light2004end,brewer,crameri2018geodynamic,mycarta}.
 Colormaps are plotted by the lightness as calculated in perceptually-uniform colorspace \textit{CAM02-UCS} in order to demonstrate this important property for making a colormap choice.
 
-\begin{figure*}
-	\includegraphics[width=\textwidth]{figures/allmaps.pdf}
-	\caption{}
-	\label{fig:allmaps}
-\end{figure*}
 % colormaps
-\nocite{hunter2007matplotlib}
-\nocite{nunez2018optimizing}
-\nocite{cubehelix}
-\nocite{twilight}
-\nocite{matlab}
-\nocite{mycarta}
-\nocite{light2004end}
-\nocite{brewer}
-\nocite{crameri2018geodynamic}
 
@@
 
-Colormaps have fundamental control over interpretation of data, and thus should be given some attention.
-Background information based on human perception of color as well as some basic design advice has been presented to guide selections.
+Colormaps have fundamental control over interpretation of data, and thus should be given careful attention. Human perception of color as well as basic design advice should guide selections.
+
+The notebook used to create the base figures in this paper can be found at \url{https://github.com/kthyng/cise-colormaps}.
 
@@
 
-The notebook used to create the base figures in this paper can be found at \url{https://github.com/kthyng/cise-colormaps}.
-
-Plots were created using the following Python packages: matplotlib \citep{hunter2007matplotlib}, Jupyter \citep{kluyver2016jupyter}, numpy \citep{oliphant2006guide,walt2011numpy},
-xarray \citep{hoyer2017xarray},
-pandas \citep{mckinney-proc-scipy-2010}, cartopy \citep{Cartopy}, and colorspacious \citep{colorspacious} along with the colormaps packages cmocean, colorcet \citep{colorcet}, and palettable \citep{palettable}.
+Plots were created using the following Python packages: matplotlib, Jupyter, numpy, xarray, pandas, cartopy, and colorspacious along with the colormaps packages cmocean, colorcet, and palettable.
 
@@
 
+\begin{figure*}
+	\includegraphics[width=\textwidth]{figures/motivation.pdf}
+	\caption{}
+	\label{fig:motivation}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{figures/evaluation.pdf}
+	\caption{}
+	\label{fig:evaluation}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{figures/howtochoose.pdf}
+	\caption{}
+	\label{fig:howtochoose}
+\end{figure*}
+
+\begin{figure*}
+	\includegraphics[width=\textwidth]{figures/allmaps.pdf}
+	\caption{}
+	\label{fig:allmaps}
+\end{figure*}
+
 \end{document}

### 2020-06-30 Thyng2020/paper.tex

@@
 
-Plots were created using the following Python packages: matplotlib, Jupyter, numpy, xarray, pandas, cartopy, and colorspacious along with the colormaps packages cmocean, colorcet, and palettable.
+Plots were created using the following Python packages: Matplotlib, Jupyter, numpy, xarray, pandas, cartopy, and colorspacious along with the colormaps packages cmocean, colorcet, and palettable.
 

### 2020-08-15 CombaCOVID2020/paper.tex
adding first version of COVID-19 paper

\documentclass[10pt,journal,compsoc]{IEEEtran}

% \usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage{natbib}
%\usepackage{url}
\usepackage{hyperref}
 \usepackage{breakurl}

 \usepackage{xspace}
\newcommand{\ie}{\emph{i.e.,}\xspace}
\newcommand{\eg}{\emph{e.g.,}\xspace}
\newcommand{\ea}{\emph{et al.}\xspace}
\newcommand{\cv}{COVID-19\xspace}
\newcommand{\tool}{COVID-19 Analysis Tools\xspace}

% \newcommand{\cm}[1]{\textbf{#1}}

\begin{document}

\title{The Importance of Data Visualization for the Understanding of \cv}

\author{%
	João L. D. Comba
	\IEEEcompsocitemizethanks{
	\IEEEcompsocthanksitem João L. D. Comba is at Instituto de Informática, UFRGS, Brazil.\protect
}}

\IEEEtitleabstractindextext{%
\begin{abstract}
Visualization techniques have been front-and-center in the efforts to communicate the science around COVID-19 broadly. In this paper I summarize and illustrate with examples how visualization can help understand different aspects of the pandemic.  \end{abstract}
}

% make the title area
\maketitle

%Up to 2,500-3,000 words in length, including the abstract, references, bios, figures, and all other text in the article.  When counting words, note that tables and figures should be counted as 250 words each.

%\section*{First Approaches to visualize \cv data}

The deadly impact of the Coronavirus pandemic (\cv) in the world is driving a massive amount of research that aims at understanding the various characteristics of the pandemic. While there is no vaccine, considerable effort has been devoted to understanding the spread of the disease in different places in the world. The speed with which the disease has spread throughout the world demands agile solutions to understand and estimate the disease progression.

Interactive \emph{dashboards} with several charts surfaced in different formats to offer concise ways to express the pandemic's growth. Fig.~\ref{fig:dashboards} illustrates some examples. The dashboard developed by the Johns Hopkins University (JHU)~\cite{dashboardJHU} was the first to track and display information on case and death totals for different countries and states in the United States. Along with lists of total counts and histograms, a \emph{bubble map} composed of circles of different radii allows a visual inspection of how serious the pandemic is around the world. Interesting plots were created on news vehicles such as the New York Times (NYT)~\cite{NYT,WP} and the Washington Post~\cite{WP}. For example, NYT used \emph{Choropleth maps}, a representation where geographical regions (countries or states) are mapped to colors associated with a measurement for that region (\eg number of cases). Such representation is useful to communicate trends, such as the average daily counts for the past week. Similarly, they display the time series evolution for each region using \emph{line heatmaps}, where daily values are mapped to colors and displayed in a row. 
We used both Choropleth maps and line heatmaps in the development of dashboard specially designed for Brazil. Using a focus-plus-context interface with coordinated views, the user can inspect the context using two Choropleth maps (one for the states and a second for a state selected), as well as details using two \emph{matrix heatmaps} (collection of stacked line heatmaps) for states and cities. The tool displays daily or cumulative data, absolute or relative to population, and supports filtering by a time interval.

%
A vast collection of community-developed dashboards and interactive tools about \cv are available. Good starting places to lool is the data hub hosted by Tableau and the top 100 R resources organized by Soetewey~\cite{R-resources}. In-depth analysis is available at sites such as \emph{Our World in Data}\cite{OWID},  \emph{Bing}~\cite{BING} and the \emph{COVID Tracking Project}~\cite{COVID-TRACKING}, among others. After developing the Brazilian dashboard, we devoted our efforts to create a set of tools to compared the spread of \cv data in different regions in the world~\cite{covid19-tools}. We collected data from more than 6000 locations in the world. Our interface has different charts that support mixing locations from different places in a single chart. Since the pandemic is at different stages in the world, we allow the user to align the time-series of data by a certain data the series passes a given threshold (\eg after 100 cases). Such representation is useful to observe when different locations passed through specific checkpoints (top of Fig.~\ref{fig:covid19-tool}).

While our initial charts support the comparison of different regions in the world, it does require that the user drives the process of choosing the regions to inspect. From the beginning, we felt the need to have an automatic tool that could, given a region of interest, return the closest regions given a similarity function. Looking at the disease's spread in distinct places, but with similar growth patterns, can be useful to predict behaviors. Thus, we developed a search engine to support queries using different similarity functions. For example, at the bottom of Fig.~\ref{fig:covid19-tool}, we show the results of searching for regions similar to Italy concerning the number of deaths. The results are listed in a ranking, with pairwise comparisons that detail attributes of the different locations and evolution charts, aligned by the day of the first death. We observe the similarities in the evolution in the number of deaths and cases for Italy, France, Spain and the United Kingdom. While using the tool, we also saw similarities among cities from Brazil and the United States, both countries with large \cv numbers.

The examples so far give a glimpse of how data visualization can help in the understanding of \cv. Fig.~\ref{fig:applications} illustrates other applications where data visualization can help. 
The first example shows how \emph{multi-dimensional projections} and \emph{network visualizations} can help the \emph{literature exploration} of papers that describe CoronaVirus clinical trials. As new research about \cv is published, there is a great need to review up-to-date literature and treatments conveniently. 
\emph{Contact tracing} is another application that relies on graph visualization to trace the network of \cv contagious people, an effort essential to control the dissemination of the disease and essential for relaxing social distancing measures. Graph visualization is also important in for \emph{social media spread and fact checking}. With many people at home, social networks are playing a significant role in people's lives these days. Unfortunately, the dissemination of fake news and automated posting from robots is also rising. Fact-checking over the propagation network can help identify misleading information and patterns of dissemination. The fourth and final example highlights the importance of data visualization in the analysis of scientific simulations. It shows the visualization of simulating the transport and spread of Corona Virus in closed spaces~\cite{VUORINEN2020}, which shows how an infected person can disseminate the virus indoors.

Many other examples include data visualization in the analysis of \cv data, and many more is surfacing every day. We hope this summary highlights interesting examples, give pointers to other references, and motivate people to pursue other applications.

\bibliographystyle{IEEEtran}
\bibliography{refs}

\bigskip

\textbf{João L. D. Comba} is a full professor of Computer Science at Instituto de Informática - UFRGS - Brazil. His research interests include visualization, visual analytics, data science, geometric algorithms and spatial data structures. He received a Ph.D. in computer science from the Stanford University, USA. Contact him at comba@inf.ufrgs.br.

\begin{figure*}
	\includegraphics[width=0.95\textwidth]{figures/jhu-dashboard2.pdf}
	\includegraphics[width=0.95\textwidth]{figures/ny-times.pdf}
	\includegraphics[width=0.95\textwidth]{figures/brazil-dashboard.pdf}
	\caption{Dashboads and interactive tools to analyse COVID-19 data: JHU Dashboard, The New York Times charts and INF-UFRGS Brazil Dashboard.}
	\label{fig:dashboards}
\end{figure*}

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/chart2.pdf}
	\includegraphics[width=\textwidth]{figures/similarity1.pdf}
	\includegraphics[width=\textwidth]{figures/similarity2.pdf}
	\caption{(top) Heatmap matrices are useful for comparing time series such as the total deaths for different countries. Columns can be aligned by the first date after reaching a certain threshold, which allows to compare when countries passed through specific checkpoints. (bottom) Searching for places with similar timelines of deaths to Italy.}
	\label{fig:covid19-tool}
\end{figure*}

\begin{figure*}
	\includegraphics[width=\textwidth]{figures/applications1.pdf}
	\includegraphics[width=\linewidth]{figures/applications2.pdf}
	\includegraphics[width=\linewidth]{figures/applications3.pdf}
	\caption{Applications: Literature exploration, Contact Tracing, Spread of claims and fact checking, and simulation of the transport and spread of Corona virus}
	\label{fig:applications}
\end{figure*}

\end{document}

### 2020-08-17 CombaCOVID2020/paper.tex

@@
+#
+
 \documentclass[10pt,journal,compsoc]{IEEEtran}
@@
 
-\title{The Importance of Data Visualization for the Understanding of \cv}
+\title{Data Visualization for the Understanding of \cv}
 
@@
 \begin{abstract}
-Visualization techniques have been front-and-center in the efforts to communicate the science around COVID-19 broadly.
+Visualization techniques have been front-and-center in the efforts to communicate the science around COVID-19 to the very broad audience of policy makers, scientists, healthcare providers, and the general public.
 In this paper I summarize and illustrate with examples how visualization can help understand different aspects of the pandemic.
@@
 
-The deadly impact of the Coronavirus pandemic (\cv) in the world is driving a massive amount of research that aims at understanding the various characteristics of the pandemic.
+The deadly impact of \cv is driving a massive amount of research that aims at understanding the various characteristics of the pandemic.
 While there is no vaccine, considerable effort has been devoted to understanding the spread of the disease in different places in the world.
@@
 Fig.~\ref{fig:dashboards} illustrates some examples.
-The dashboard developed by the Johns Hopkins University (JHU)~\cite{dashboardJHU} was the first to track and display information on case and death totals for different countries and states in the United States.
+The dashboard developed by Johns Hopkins University (JHU)~\cite{dashboardJHU} was the first to track and display information on case and death totals for different countries and states in the United States.
 Along with lists of total counts and histograms, a \emph{bubble map} composed of circles of different radii allows a visual inspection of how serious the pandemic is around the world.
-Interesting plots were created on news vehicles such as the New York Times (NYT)~\cite{NYT,WP} and the Washington Post~\cite{WP}.
+Interesting plots were created on news outlets such as the New York Times (NYT)~\cite{NYT,WP} and the Washington Post~\cite{WP}.
 For example, NYT used \emph{Choropleth maps}, a representation where geographical regions (countries or states) are mapped to colors associated with a measurement for that region (\eg number of cases).
-Such representation is useful to communicate trends, such as the average daily counts for the past week.
+This representation is useful to communicate trends, such as the average daily counts for the past week.
 Similarly, they display the time series evolution for each region using \emph{line heatmaps}, where daily values are mapped to colors and displayed in a row.
-We used both Choropleth maps and line heatmaps in the development of dashboard specially designed for Brazil.
+Our work (described in more detail below) used both Choropleth maps and line heatmaps in the development of dashboard specially designed for Brazil.
 Using a focus-plus-context interface with coordinated views, the user can inspect the context using two Choropleth maps (one for the states and a second for a state selected), as well as details using two \emph{matrix heatmaps} (collection of stacked line heatmaps) for states and cities.
@@
 A vast collection of community-developed dashboards and interactive tools about \cv are available.
-Good starting places to lool is the data hub hosted by Tableau and the top 100 R resources organized by Soetewey~\cite{R-resources}.
+Good starting places to look are the data hub hosted by Tableau and the top 100 R resources organized by Soetewey~\cite{R-resources}.
 In-depth analysis is available at sites such as \emph{Our World in Data}\cite{OWID},  \emph{Bing}~\cite{BING} and the \emph{COVID Tracking Project}~\cite{COVID-TRACKING}, among others.
-After developing the Brazilian dashboard, we devoted our efforts to create a set of tools to compared the spread of \cv data in different regions in the world~\cite{covid19-tools}.
-We collected data from more than 6000 locations in the world.
-Our interface has different charts that support mixing locations from different places in a single chart.
+After developing the Brazilian dashboard, we devoted our efforts to create a set of tools to compare the spread of \cv data in different regions of the world~\cite{covid19-tools}.
+We collected data from more than 6,000 locations in the world, and our interface has different charts that support visualizing multiple locations in a single chart.
 Since the pandemic is at different stages in the world, we allow the user to align the time-series of data by a certain data the series passes a given threshold (\eg after 100 cases).
-Such representation is useful to observe when different locations passed through specific checkpoints (top of Fig.~\ref{fig:covid19-tool}).
+This representation is useful to observe when different locations passed through specific checkpoints (top of Fig.~\ref{fig:covid19-tool}).
 
-While our initial charts support the comparison of different regions in the world, it does require that the user drives the process of choosing the regions to inspect.
+While our initial charts support the comparison of different regions, the tool required the user to drive the process of choosing the regions to compare.
 From the beginning, we felt the need to have an automatic tool that could, given a region of interest, return the closest regions given a similarity function.
@@
 The examples so far give a glimpse of how data visualization can help in the understanding of \cv. Fig.~\ref{fig:applications} illustrates other applications where data visualization can help.
-The first example shows how \emph{multi-dimensional projections} and \emph{network visualizations} can help the \emph{literature exploration} of papers that describe CoronaVirus clinical trials.
+The first example shows how \emph{multi-dimensional projections} and \emph{network visualizations} can help the \emph{literature exploration} of papers that describe novel coronavirus research.
 As new research about \cv is published, there is a great need to review up-to-date literature and treatments conveniently.
-\emph{Contact tracing} is another application that relies on graph visualization to trace the network of \cv contagious people, an effort essential to control the dissemination of the disease and essential for relaxing social distancing measures.
+\emph{Contact tracing} is another application that relies on graph visualization to trace the network of people who may have been in contact with a \cv patient, an activity essential to control the dissemination of the disease and essential for directing social distancing regulations.
 Graph visualization is also important in for \emph{social media spread and fact checking}.
@@
 The fourth and final example highlights the importance of data visualization in the analysis of scientific simulations.
-It shows the visualization of simulating the transport and spread of Corona Virus in closed spaces~\cite{VUORINEN2020}, which shows how an infected person can disseminate the virus indoors.
+It shows the visualization of simulating the transport and spread of novel coronavirus in closed spaces~\cite{VUORINEN2020}, which shows how an infected person can disseminate the virus indoors.
 
@@
 	\includegraphics[width=\linewidth]{figures/applications3.pdf}
-	\caption{Applications: Literature exploration, Contact Tracing, Spread of claims and fact checking, and simulation of the transport and spread of Corona virus}
+	\caption{Applications: Literature exploration, contact tracing, spread of claims and fact checking, and simulation of the transport and spread of the novel coronavirus.}
 	\label{fig:applications}
